---
title: "Darwin Core mapping"
subtitle: "For: The checklist of alien seed plants of D.R. Congo, based on evidence from herbarium collections"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
#  pdf_document:
#    df_print: kable
#    number_sections: yes
#    toc: yes
#    toc_depth: 3
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/alien-csi/alien-flora-congo-checklist/tree/main/data/raw)


Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(readxl)         # To read Excel files
library(digest)         # To generate hashes
```

# Read source data

Create a data frame `input_data` from the source data:
 
```{r}
input_data <- read_excel(path = here("data", "raw", "S1 Accepted aliens revised.xlsx"),
                         col_names = FALSE) 
```

Preview data:

```{r}
input_data %>% head(n = 5)
```

# Preprocessing

## Tidy data

Remove first row, generate new headers and clean names:

```{r}
names(input_data) <- input_data[2,]

input_data <- input_data[-c(1,2),]

input_data %<>%
  remove_empty("rows") %>%       # Remove empty rows
  clean_names() 
```

## Taxon IDs

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable:

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
input_data %<>% mutate(taxon_id = paste("alien-flora-congo-checklist", "taxon", vdigest(accepted_name, algo="md5"), sep=":"))
```

Preview data:

```{r}
input_data %>% head()
```


# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml):

### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "Botanic Garden Meise")
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "The checklist of alien seed plants of D.R. Congo, based on evidence from herbarium collections")
```

### taxonID   

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = accepted_name)
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = "Plantae")
```

### family

```{r}
taxon %<>% mutate(dwc_family = family)
```

### taxonRank

```{r}
taxon %<>% mutate(dwc_taxonRank = "species")
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = "ICN")
```

## Distribution extension

```{r start_distribution}
distribution <- input_data
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml):

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id)
```

### locality

```{r}
distribution %<>% mutate(dwc_locality = "the Democratic Republic of the Congo")
```

### countryCode

```{r}
distribution %<>% mutate(dwc_countryCode = "CD")
```

### establishmentMeans

Values for `degreeOfEstablishment` are contained in `proposed_status`

```{r}
distribution %<>% mutate(dwc_establishmentMeans = case_when(
  proposed_status == "Naturalised, cryptogenic" |
    proposed_status == "Naturalised, cryptogenic, invasive" ~ "uncertain",
  TRUE ~ "introduced"))
```

### degreeOfEstablishment

Values for `degreeOfEstablishment` are contained in `proposed_status`

```{r}
distribution %>% 
  group_by(proposed_status) %>% 
  summarize(records = n())
```

```{r}
distribution %<>% mutate(dwc_degreeOfEstablishment = case_when(
  proposed_status == "casual" |
    proposed_status == "Casual" ~ "casual",
  proposed_status == "Naturalised" |
    proposed_status == "Naturalised, cryptogenic" ~ "established",
  proposed_status == "Naturalised, cryptogenic, invasive" |
    proposed_status == "Naturalised, invasive" ~ "invasive"))
```


### eventDate

Screen `earliest_record` and `latest_record`

```{r}
distribution %>% 
  group_by(earliest_record, latest_record) %>% 
  summarize(records = n())
```

```{r}
distribution %<>% 
  mutate(dwc_eventDate = case_when(
    is.na(earliest_record) ~ latest_record,
    earliest_record == "s.d." & latest_record == "s.d." ~ "",
    is.na(latest_record) ~ earliest_record,
    latest_record == "s.d." & earliest_record == "1937" ~ earliest_record,
    is.na(earliest_record) & is.na(latest_record) ~ "",
    TRUE ~ paste(earliest_record, "/", latest_record)
  ))
  
```


## Description extension

In the description extension we want to include several important characteristics (hereafter reffered to as descriptors) about the species:

- Native range
- Life form

A single taxon can have multiple descriptions of the same type (e.g. multiple native ranges), expressed as multiple rows in the description extension.

For each descriptor, we create a separate dataframe to process the specific information. We always specify _which descriptor_ we map (`type` column) and its _specific content_ (`description` column). After the mapping of these Darwin Core terms `type` and `value`, we merge the dataframes to generate one single description extension. We then continue the mapping process by adding the other Darwin Core terms (which content is independent of the type of descriptor, such as `language`).

### native range
 
Create separate dataframe:

```{r}
native_range <- input_data
```

Screen `continent_of_origin`:

```{r}
native_range %>% 
  group_by(continent_of_origin) %>% 
  summarize(records = n())
```

Clean up `continent_of_origin`

```{r}
native_range <-
  native_range %>% 
    mutate(continent_of_origin = str_trim(continent_of_origin)) %>% 
    mutate(continent_of_origin = recode(continent_of_origin,
                                        "Hybr" = "Multiple"
    ))
```

Create `description` from `continent_of_origin`:

```{r}
native_range <-
  native_range %>% 
    mutate(description = continent_of_origin) %>% 
    filter(!is.na(description))
```

Create a `type` field to indicate the type of description:

```{r}
native_range %<>% mutate(type = "native range")
```

### native range
 
Create separate dataframe:

```{r}
life_form <- input_data
```

Screen `life_form`:

```{r}
life_form %>% 
  group_by(life_form) %>% 
  summarize(records = n())
```

Create `description` from `life_form`:

```{r}
life_form %<>% mutate(description = life_form)
```

Create `type`:

```{r}
life_form %<>% mutate(type = "life form")
```

Union native range and life form:

```{r}
description <- bind_rows(native_range, life_form)
```

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml):

### taxonID

```{r}
description %<>% mutate(dwc_taxonID = taxon_id)
```

### description

```{r}
description %<>% mutate(dwc_description = description)
```

### type

```{r}
description %<>% mutate(dwc_type = type)
```

### language

```{r}
description %<>% mutate(dwc_language = "en")
```

# Post-processing

Remove the original columns for each of the generated files:

```{r}
taxon %<>% select(starts_with("dwc_")) 
distribution %<>% select(starts_with("dwc_")) 
species_profile %<>% select(starts_with("dwc_")) 
description %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
colnames(species_profile) <- str_remove(colnames(species_profile), "dwc_")
colnames(description) <- str_remove(colnames(description), "dwc_")
```

Sort on `taxonID` to group description information per taxon:

```{r}
taxon %<>% arrange(taxonID)
distribution %<>% arrange(taxonID)
species_profile %<>% arrange(taxonID)
description %<>% arrange(taxonID)
```

Preview taxon core:

```{r}
taxon %>% head(10)
```

Preview distribution extension:

```{r}
distribution %>% head(10)
```

Preview species profile extension:

```{r}
species_profile %>% head(10)
```

Preview description extension:

```{r}
description %>% head(10)
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```


